---
#########################################
# options for knitting a single chapter #
#########################################
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
    citation_package: biblatex
  bookdown::html_document2: default
  bookdown::word_document2: default
documentclass: book
bibliography: references.bib
---

```{r echo = FALSE}
options(scipen = 100)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

# Introduction
\adjustmtc
\markboth{Introduction}{}
<!-- For PDF output, include these two LaTeX commands after unnumbered chapter headings, otherwise the mini table of contents and the running header will show the previous chapter -->

```{r}
custom_numbers_to_words <- function(number) {
  powers <- c(3, 6, 9, 12)
  
  if (number == 0) {
    return("0")
  }
  
  sign <- ifelse(number < 0, -1, 1)
  number <- abs(number)
  
  for (power in powers) {
    scaled_number <- number / 10^power
    if (scaled_number >= 1 && scaled_number < 1000) {
      suffix <- gsub("^one\\s", "", xfun::numbers_to_words(10^power))
      return(paste0(sign * scaled_number, " ", suffix))
    }
  }
  
  return("Number out of range.")
}

esa_plhiv2022 <- 20.8 * 1e6
wca_plhiv2022 <- 4.8 * 1e6
ssa_plhiv2022 <- esa_plhiv2022 + wca_plhiv2022

plhiv2022 <- 39 * 1e6

ssa_pop2022 <- 1.2 * 1e9 #' From the world bank
```

This thesis is about applied and methodological Bayesian statistics.
The sense in which it is Bayesian is that I use probability models to arrive at conclusions based on data.
The sense in which it is applied and methodological is that I am concerned with real world questions and the means to answer them.

The real world questions relate to surveillance of the HIV epidemic in sub-Saharan Africa, a region it is estimated that `r round(ssa_plhiv2022 / plhiv2022 * 100)`% of the total `r custom_numbers_to_words(plhiv2022)` people living with HIV in the world live [@unaids2022global].
Quantifying the epidemic using statistics is an important part of the public health response, and the path towards disease control and elimination. 

The data in this thesis appertain to people's answers to survey questions or use of healthcare facilities.
An important feature of the data is its location in space and time.
Spatio-temporal data, while diverse, have distinctive commonalities which make its collective study valuable.

Computation is an essential part of modern statistical practice.
Each project chapter is accompanied by code, hosted on GitHub.

## Chapter overview

* Chapter \@ref(hiv-aids): I begin by providing background on the HIV/AIDS epidemic.
* Chapter \@ref(bayes-st): I then introduce the statistical concepts and notation used throughout the thesis.
* Chapter \@ref(beyond-borders): The prevailing model for spatial structure used in small-area estimation [@besag1991bayesian] was designed with analysis of a grid of pixels in mind.
In disease mapping, we work using the districts of a country, which are not a grid.
I evaluate the practical consequences of this this concern [@howes2023beyond].
* Chapter \@ref(multi-agyw): Adolescent girls and young women are a demographic group at disproportionate risk of HIV infection.
The Global AIDS Strategy suggests prioritising interventions on the basis of behaviour to prevent the most new infections using available resources.
I estimate the size of behavioural risk groups across priority countries to enable implementation of this strategy, and assess the potential benefits in terms of numbers of new infections prevented [@howes2023spatio].
* Chapter \@ref(naomi-aghq): The Naomi small-area estimation model [@eaton2021naomi] is used by countries to estimate district-level HIV indicators.
Motivated by this model, I develop an approximate Bayesian inference method combining adaptive Gauss-Hermite quadrature with principal components analysis [@howes2023fast].
I apply the method to data from Malawi, and analyse the consequences of inference method choice for policy relevant outcomes.
Further, I open the door to a new class of fast, flexible, and accurate Bayesian inference algorithms.
* Chapter \@ref(conclusions): I discuss avenues for future work, and my conclusions regarding the research.

Figure \@ref(fig:chapter-flowchart) gives the dependency structure for the above chapters.
Though the presented order is recommended, Chapter \@ref(beyond-borders), Chapter \@ref(multi-agyw) and Chapter \@ref(naomi-aghq) may be read in any order.

```{r chapter-flowchart, fig.cap="Flowchart presenting the recommended chapter dependencies for this thesis. The dashed line represents a recommended, but not required dependency."}
knitr::include_graphics("figures/chapter-flowchart.pdf")
```
