---
#########################################
# options for knitting a single chapter #
#########################################
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
    citation_package: biblatex
  bookdown::html_document2: default
  bookdown::word_document2: default
documentclass: book
bibliography: references.bib
---

```{r echo = FALSE}
options(scipen = 100)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

# Background {#background}
\adjustmtc
\markboth{Background}{}
<!-- For PDF output, include these two LaTeX commands after unnumbered chapter headings, otherwise the mini table of contents and the running header will show the previous chapter -->

## The HIV/AIDS epidemic

```{r}
plhiv2022 <- 38000000
deaths2022 <- 700000
infections2022 <- 1700000
```

<!-- The extent of HIV as a problem -->
According to latest estimates, in 2022 there were `r xfun::numbers_to_words(plhiv2022)` people living with HIV, `r xfun::numbers_to_words(deaths2022)` AIDS-related deaths, and `r xfun::numbers_to_words(infections2022)` people newly infected with HIV.

<!-- Uneven burden -->
Disease burden is unevenly distributed in space, and across communities and individuals.
The region most affected is Sub-Saharan Africa.
Key populations include men who have sex with men, female sex workers, people who inject drugs, transgender people, incarcerated people.
Large demographic groups at higher risk include adolescent girls and young women.
Fortunately, progress is being made.

<!-- Possible interventions to help -->
Key interventions are ART, condoms, PrEP and PEP, education, economic empowerment, VMMC.

<!-- Prioritisation of interventions  -->
Methods for prevention prioritisation include geographic, demographic, key population services, risk screening, individual-level risk characteristics.
Are there differences in effectiveness of treatments for different groups.

<!-- What is surveillance and how can it be used to meet these needs -->
Surveillance is used is conducted to track epidemic trends, identify at-risk populations, find drivers of transmission, and evaluate the impact of prevention and treatment programs.
Key HIV indicators are HIV prevalence, HIV incidence, coverage of ART and other interventions.

<!-- Why is surveillance difficult -->
There are significant data related difficulties associated with furnishing these estimates.
These include sparsity in space and time, survey bias, conflicting information sources, hard to reach populations, changing demographies.
These data limitations foreground the importance of synthesising multiple sources of information to obtain estimates.
Doing so increases the difficulty and complexity of the statistical modelling required.

<!-- Future trends in surveillance -->
Aims for HIV response going forward, and surveillance capabilities are needed to meet them.
Phasing out of nationally-representative household surveys for HIV.

## Bayesian spatio-temporal statistics

### Bayesian statistics

Bayesian statistics is a paradigm for learning from data.
In this section I will provide a brief overview.
For a more complete introduction, I recommend @mcelreath2020statistical or @gelman2013bayesian.

At it's best, the Bayesian framework allows the analyst focus their attention on modelling the data at hand.
In particular, this is achieved by the construction of a generative model for the observed data $y$ together with parameters $\vartheta$
\begin{equation}
(y, \vartheta) \sim p(y, \vartheta).
\end{equation}
Given this generative model, computation of the posterior distribution
\begin{equation}
p(\vartheta \, | \, y) = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}, \label{eq:posterior}
\end{equation}
usually proceeds using approximate Bayesian inference methods, where $p(y \, | \, \vartheta)$ is the likelihood, $p(y \, | \, \vartheta)$ is the prior, and $p(\vartheta)$ is the evidence.

Markov chain Monte Carlo (MCMC) is the most popular approach for computing Equation \ref{eq:posterior}, and proceeds by simulating samples from a Markov chain with stationary distribution equal to the distribution of interest.
Variational Bayes approaches assume the posterior distribution belongs to some class and use optimisation to choose the best member of that class.
Particular properties of spatio-temporal models make integrated nested Laplace approximations, if feasible, often the best option
Empirical Bayes approaches, like Template Model Builder [@osgoodzimmerman2021statistical].

### Spatio-temporal statistics

In spatio-temporal statistics the data we observe are indexed by spatial or temporal location.
Commonly used independent and identically distributed (IID) assumptions on observations are rarely suitable in this setting because we expect there to be spatio-temporal correlation structure.

Split the parameters $\vartheta = (x, \theta)$.
Call $x$ the latent field.
Call $\theta$ the hyperparameters.
Often, the latent field is assumed to be jointly multivariate Gaussian.

### Latent Gaussian and extended latent Gaussian models

Latent Gaussian models [@rue2009approximate] are of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $[n] = \{1, \ldots, n\}$.
The response variable is $y = (y)_{i \in [n]}$ with likelihood $p(y \, | \, \eta, \theta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \theta_1)$, where $\eta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\theta_1 \in \mathbb{R}^s$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $x \in \mathbb{R}^N$ called the latent field such that $x \sim \mathcal{N}(0, Q(\theta_2)^{-1})$ where $\theta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\theta = (\theta_1, \theta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\theta)$.
Common examples of latent Gaussian models include the following.

Many of the leading-edge models used in small-area estimation fall outside the latent Gaussian model class.
Examples include disaggregation models, evidence synthesis models [@eaton2019joint; @eaton2021naomi], attendance models, risk group models.
<!-- There are other complex models from [ecology](https://www.bioss.ac.uk/rsse/2013/September2013slides-Illian.pdf) that can't currently be fit using `R-INLA` -->
However, many of these models do fit into the class of extended latent Gaussian models [@stringer2021fast].
By allowing many-to-one link functions, extended latent Gaussian models facilitate modelling of non-linearities.
<!-- Strictly speaking, many-to-one is not an issue for `R-INLA`, the latent field is implemented as a concatenation of many vectors already e.g. for $\eta_i = \beta_0 + \phi_i$ with $i = 1, \ldots, n$ the latent field is $(\eta_1, \ldots, \eta_n, \beta_0, \phi_1, \ldots, \phi_n)^\top$ of dimension $2n + 1$ -->

<!-- Adaptive Gauss Hermite quadrature (AGHQ), like the central composite design (CCD) and grid strategies, is one way to choose the hyper-parameter integration points in the integrated nested Laplace approximation (INLA) -->
<!-- Finn Lindgren is working on a method for non-linear predictors, called the [iterative INLA method](https://github.com/inlabru-org/inlabru/blob/55896f10d563c14e34cab577b29b733aac051f86/vignettes/method.Rmd). More [slides](https://informatique-mia.inrae.fr/reseau-resste/sites/default/files/2020-09/slides-Lindgren_Avignon2018.pdf) here -->
<!-- Thesis work of [Follestad](https://cds.cern.ch/record/639625/files/sis-2003-305.pdf) that stayed as a preprint -->
<!-- How does the ecological fallacy relate to aggregated output models? -->

### Survey methods
